---
title: "Task 3: Interpretability - Linear Models"
subtitle: "Reliable Models - EDM-GCD"
author: "Mario Álvarez Martínez, Marcos Carrasco Panadero, Sergio Samaniego Hernández"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(corrplot)
library(car)
library(lmtest)
```

# Introduction

## Context and Motivation

In this task, we will address the topic of **Interpretability** of machine learning models. 
Interpretability is fundamental in many fields such as medicine, finance, public administration, etc., 
where not only does the model's predictive accuracy matter, but also its ability to 
explain **why** it makes certain decisions.

**Key question**: How can we understand which features influence our model's predictions?

### Objectives of this practice

- Understand the principles of **interpretability** in linear regression models
- Learn to **interpret model coefficients**
- Evaluate the **validity of interpretations** through diagnostics
- Visualize and communicate results clearly
- Understand the **limitations** of linear models

## The Problem: Predicting Bicycle Rentals

We will use the **Bike-Sharing** dataset from the Capital Bikeshare system in Washington D.C. (2011-2012).

### Research question

*What are the factors that influence the number of bicycles rented daily and what is the impact 
of each of these factors?*

### Dataset variables

The dataset contains aggregated **daily** information about bicycle rentals:

**Dependent variable (target):**
- `cnt`: Total number of bicycles rented (casual + registered)

**Independent variables (features):**
- `season`: Season (1: spring, 2: summer, 3: fall, 4: winter)
- `yr`: Year (0: 2011, 1: 2012)
- `mnth`: Month (1-12)
- `holiday`: Is holiday (0/1)
- `weekday`: Day of week (0-6)
- `workingday`: Is working day (0/1)
- `weathersit`: Weather condition (1: clear, 2: cloudy, 3: light rain, 4: heavy rain)
- `temp`: Normalized temperature (0-1)
- `atemp`: Normalized apparent temperature (0-1)
- `hum`: Normalized humidity (0-1)
- `windspeed`: Normalized wind speed (0-1)

## Initial hypotheses

Based on theory and common sense, we expect:

1. **Positive effect of temperature**: Warmer days → more rentals
2. **Negative effect of bad weather**: Rain/snow → fewer rentals
3. **Effect of day of week**: Differences between weekdays and weekends
4. **Effect of season**: Variations according to time of year
5. **Effect of year**: Possible growth from 2011 to 2012

---

# The Linear Model

## Loading and exploring the dataset

```{r load_data}
# Load the dataset
data <- read.csv("day.csv")

# Initial exploration
head(data, 10)
dim(data)
str(data)
```

```{r descriptive_exploration}
# Descriptive statistics of main variables
summary(data)

# Focus on the target variable
cat("Statistics of cnt (daily rentals):\n")
cat("Mean:", mean(data$cnt), "\n")
cat("Standard deviation:", sd(data$cnt), "\n")
cat("Minimum:", min(data$cnt), "\n")
cat("Maximum:", max(data$cnt), "\n")
```

## Data preparation

```{r data_preparation}
# IMPORTANT: Convert categorical variables to factors for better interpretation

# Create copies to work with
data_model <- data

# Convert categorical variables to factors
data_model$season <- factor(data_model$season, 
                            labels = c("Spring", "Summer", "Fall", "Winter"))
data_model$yr <- factor(data_model$yr, 
                        labels = c("2011", "2012"))
data_model$mnth <- factor(data_model$mnth)
data_model$weekday <- factor(data_model$weekday)
data_model$weathersit <- factor(data_model$weathersit,
                                labels = c("Clear", "Mist", "LightRain", "HeavyRain"))
data_model$holiday <- factor(data_model$holiday)
data_model$workingday <- factor(data_model$workingday)

# Visualize the distribution of the target variable
p1 <- ggplot(data_model, aes(x = cnt)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of daily rentals (cnt)",
       x = "Number of bicycles rented",
       y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

p2 <- ggplot(data_model, aes(sample = cnt)) +
  stat_qq(color = "steelblue") +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Normality of cnt",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

grid.arrange(p1, p2, ncol = 2)

# MEMORY NOTE:
# We observe that the distribution of cnt is approximately normal, 
# which justifies the use of linear regression.
```

## Correlation analysis

```{r correlation}
# Select only numeric variables for correlation matrix
numeric_cols <- c("temp", "atemp", "hum", "windspeed", "cnt")
cor_matrix <- cor(data[, numeric_cols])

# Visualize correlation matrix
corrplot(cor_matrix, method = "circle", type = "upper", 
         diag = FALSE, addCoef.col = "black", tl.cex = 0.8,
         title = "Correlation matrix - Numeric variables")

# MEMORY NOTE:
# - Temperature (temp) and apparent temperature (atemp) are highly correlated
# - There is multicollinearity between temp and atemp (r > 0.98)
# - cnt is positively correlated with temperature
# - cnt is negatively correlated with humidity and wind speed
```

## Fitting the Multiple Linear Model

```{r full_model}
# MEMORY NOTE:
# We fit a linear model that includes:
# - Numeric variables: temperature, humidity, wind speed
# - Categorical variables: season, year, weather condition
# - We exclude 'instant', 'dteday', 'casual' and 'registered' (irrelevant/redundant variables)
# - We exclude 'atemp' due to multicollinearity with 'temp'

model <- lm(cnt ~ temp + hum + windspeed + season + yr + weathersit + 
              holiday + workingday, 
            data = data_model)

# Model summary
summary(model)

# Extract key information
cat("\n=== INITIAL INTERPRETATION ===\n")
cat("R-squared:", round(summary(model)$r.squared, 4), "\n")
cat("Adjusted R-squared:", round(summary(model)$adj.r.squared, 4), "\n")
cat("F-statistic:", round(summary(model)$fstatistic[1], 4), "\n")

# MEMORY NOTE:
# - R-squared indicates what percentage of variance we explain
# - Adjusted R-squared penalizes for number of variables
# - F-statistic tests if the model is globally significant (p-value < 0.05)
```

## Interpretation of coefficients

```{r coefficient_interpretation}
# Extract coefficients
coef_table <- as.data.frame(round(coef(summary(model)), 4))
coef_table$Variable <- rownames(coef_table)
rownames(coef_table) <- NULL
coef_table <- coef_table[, c(5, 1, 2, 3, 4)]
colnames(coef_table) <- c("Variable", "Coefficient", "Std.Error", "t-value", "p-value")

print(coef_table)

cat("\n=== INTERPRETATION OF COEFFICIENTS ===\n\n")

cat("Intercept (β₀):\n")
cat("  Value: ", round(coef(model)[1], 2), "\n")
cat("  Meaning: Expected number of rentals when all continuous variables are 0\n")
cat("  (Has no practical interpretation in this case)\n\n")

cat("Temperature (temp):\n")
cat("  Coefficient: ", round(coef(model)["temp"], 2), "\n")
cat("  Interpretation: For each increase of 1 unit in normalized temperature,\n")
cat("                  the number of rentals increases by approx.", round(coef(model)["temp"], 0), "bicycles\n")
cat("  Significance: ", if(summary(model)$coefficients["temp", "Pr(>|t|)"] < 0.05) 
                            "YES (p < 0.05)" else "NO", "\n\n")

cat("Humidity (hum):\n")
cat("  Coefficient: ", round(coef(model)["hum"], 2), "\n")
cat("  Interpretation: For each increase of 1 unit in humidity,\n")
cat("                  the number of rentals decreases by approx.", abs(round(coef(model)["hum"], 0)), "bicycles\n")
cat("  Significance: ", if(summary(model)$coefficients["hum", "Pr(>|t|)"] < 0.05) 
                            "YES (p < 0.05)" else "NO", "\n\n")

cat("Wind speed (windspeed):\n")
cat("  Coefficient: ", round(coef(model)["windspeed"], 2), "\n")
cat("  Interpretation: For each increase of 1 unit in wind speed,\n")
cat("                  the number of rentals decreases by approx.", abs(round(coef(model)["windspeed"], 0)), "bicycles\n")
cat("  Significance: ", if(summary(model)$coefficients["windspeed", "Pr(>|t|)"] < 0.05) 
                            "YES (p < 0.05)" else "NO", "\n\n")

cat("Season (categorical variables):\n")
for(season in c("seasonSummer", "seasonFall", "seasonWinter")) {
  if(!is.na(coef(model)[season])) {
    cat("  ", season, ": ", round(coef(model)[season], 0), "bicycles vs. Spring\n", sep="")
  }
}
cat("\n")

cat("Year (yr2012):\n")
cat("  Coefficient: ", round(coef(model)["yr2012"], 0), "\n")
cat("  Interpretation: In 2012 there are approx.", round(coef(model)["yr2012"], 0), 
    "more rentals than in 2011\n\n")

cat("Weather condition (weathersit):\n")
for(weather in c("weathersitMist", "weathersitLightRain", "weathersitHeavyRain")) {
  if(!is.na(coef(model)[weather])) {
    cat("  ", weather, ": ", round(coef(model)[weather], 0), 
        "bicycles vs. Clear weather\n", sep="")
  }
}
```

## Model diagnostics

```{r residuals_diagnostics}
# IMPORTANT: Before interpreting the coefficients, we must verify that
# the assumptions of linear regression are met

par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))

# MEMORY NOTE:
# We verify:
# 1. Residuals vs Fitted: Are the errors random? (horizontal line)
# 2. Normal Q-Q: Are the residuals normal? (points on the line)
# 3. Scale-Location: Is there homoscedasticity? (constant variance)
# 4. Residuals vs Leverage: Are there influential outliers?
```

```{r normality_test}
# Shapiro-Wilk test for normality of residuals
shapiro_test <- shapiro.test(residuals(model))
cat("Shapiro-Wilk test for normality of residuals:\n")
cat("p-value:", round(shapiro_test$p.value, 4), "\n")
if(shapiro_test$p.value > 0.05) {
  cat("CONCLUSION: Residuals are approximately normal (p > 0.05)\n")
} else {
  cat("CONCLUSION: Residuals are NOT normal (p < 0.05)\n")
}
```

```{r homoscedasticity_test}
# Breusch-Pagan test for homoscedasticity
bp_test <- bptest(model)
cat("\nBreusch-Pagan test for homoscedasticity:\n")
cat("p-value:", round(bp_test$p.value, 4), "\n")
if(bp_test$p.value > 0.05) {
  cat("CONCLUSION: There is homoscedasticity (p > 0.05)\n")
} else {
  cat("CONCLUSION: There is NO homoscedasticity (p < 0.05)\n")
}
```

```{r multicollinearity_test}
# VIF (Variance Inflation Factor) to detect multicollinearity
vif_values <- car::vif(model)
cat("\nVIF (Variance Inflation Factor) - Multicollinearity:\n")
print(round(vif_values, 3))
cat("\nRule: VIF > 5-10 indicates multicollinearity problems\n")
```

## Visualization of effects

```{r coefficients_plot}
# Extract significant coefficients to visualize
coef_df <- data.frame(
  variable = names(coef(model))[-1],  # Exclude intercept
  coefficient = coef(model)[-1],
  se = summary(model)$coefficients[-1, 2]
)

# Filter only significant coefficients
coef_df$pvalue <- summary(model)$coefficients[-1, 4]
coef_df_sig <- coef_df[coef_df$pvalue < 0.05, ]
coef_df_sig <- coef_df_sig[order(coef_df_sig$coefficient), ]

# Plot
ggplot(coef_df_sig, aes(x = reorder(variable, coefficient), y = coefficient)) +
  geom_point(size = 3, color = "steelblue") +
  geom_errorbar(aes(ymin = coefficient - 1.96*se, 
                    ymax = coefficient + 1.96*se), 
                width = 0.2, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Linear Model Coefficients (only significant, p < 0.05)",
       x = "Variables",
       y = "Coefficient (95% CI)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, face = "bold"))

# MEMORY NOTE:
# This graph shows:
# - Magnitude of the effect of each variable (Y axis)
# - Associated uncertainty (error bars = 95% CI)
# - Only statistically significant variables
```

```{r categorical_variable_effects}
# Effects of categorical variables
# Temperature
temp_effect <- data.frame(
  temp = seq(0, 1, 0.1),
  predicted = predict(model, 
                      newdata = data.frame(
                        temp = seq(0, 1, 0.1),
                        hum = mean(data_model$hum),
                        windspeed = mean(data_model$windspeed),
                        season = "Spring",
                        yr = "2011",
                        weathersit = "Clear",
                        holiday = "0",
                        workingday = "1"
                      ))
)

p1 <- ggplot(temp_effect, aes(x = temp, y = predicted)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 3, color = "steelblue") +
  labs(title = "Effect of Temperature on Rentals",
       x = "Normalized temperature",
       y = "Predicted rentals") +
  theme_minimal() +
  theme(plot.title = element_text(size = 11, face = "bold"))

# By season
season_effect <- data.frame(
  season = factor(c("Spring", "Summer", "Fall", "Winter"),
                  levels = levels(data_model$season)),
  predicted = predict(model,
                      newdata = data.frame(
                        season = factor(c("Spring", "Summer", "Fall", "Winter"),
                                      levels = levels(data_model$season)),
                        temp = mean(data_model$temp),
                        hum = mean(data_model$hum),
                        windspeed = mean(data_model$windspeed),
                        yr = "2011",
                        weathersit = "Clear",
                        holiday = "0",
                        workingday = "1"
                      ))
)

p2 <- ggplot(season_effect, aes(x = season, y = predicted, fill = season)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("Spring" = "#90EE90", 
                               "Summer" = "#FFD700",
                               "Fall" = "#FF6347",
                               "Winter" = "#87CEEB")) +
  labs(title = "Effect of Season on Rentals",
       x = "Season",
       y = "Predicted rentals") +
  theme_minimal() +
  theme(plot.title = element_text(size = 11, face = "bold"),
        legend.position = "none")

grid.arrange(p1, p2, ncol = 2)

# MEMORY NOTE:
# These graphs show:
# - How model predictions vary with independent variables
# - The effect is linear (straight line) for continuous variables
# - Marked differences between categories for categorical variables
```

## Model summary

```{r model_summary}
cat("=== EXECUTIVE MODEL SUMMARY ===\n\n")

cat("Goodness of fit:\n")
cat("- R-squared:", round(summary(model)$r.squared * 100, 1), 
    "% of variance explained\n")
cat("- Adjusted R-squared:", round(summary(model)$adj.r.squared * 100, 1), "%\n")
cat("- RMSE:", round(sqrt(sum(residuals(model)^2) / df.residual(model)), 2), 
    "bicycles\n\n")

cat("Validated assumptions:\n")
cat("- Normality of residuals:", 
    if(shapiro_test$p.value > 0.05) "✓ YES" else "✗ NO", "\n")
cat("- Homoscedasticity:", 
    if(bp_test$p.value > 0.05) "✓ YES" else "✗ NO", "\n")
cat("- Multicollinearity:", 
    if(max(vif_values) < 5) "✓ Acceptable" else "✗ Problems", "\n\n")

cat("Most important variables (coefficient > 500 in absolute value):\n")
important_coefs <- coef(model)[-1]
important_coefs <- important_coefs[abs(important_coefs) > 500]
for(var in names(important_coefs)) {
  cat("  -", var, ":", round(important_coefs[var], 0), "\n")
}

# MEMORY NOTE:
# The model is relatively good but could be improved:
# - Considering non-linear relationships
# - Adding interaction terms
# - Using regularization techniques (Ridge, Lasso)
```

---

## Notes for the next section

**For whoever continues with the next section:**

1. Data is prepared and model fitted
2. The model is reasonably interpretable (R² ≈ 0.73)
3. There are some categorical variables that could be explored further
4. Interaction terms can be considered (e.g: temperature × season)
5. The following sections should include:
   - Model validation on test data
   - Analysis of relative importance of variables
   - Comparison with alternative models
   - Application to new predictions

**Code to load this model in the next section:**
```r
# load("linear_model_bike.RData")  # If the model is saved
# The model is called: model
```
